Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous \emph{Arabidopsis thaliana} Incurs No Photosynthetic Advantage

### 2023-10-17 ms/ms.Rmd
added multtest results

# Summarize results on NNI
nni = single_surface_results |>
  dplyr::transmute(sig = p_value < 0.05) |>
  dplyr::summarize(n1 = sum(sig), n2 = sum(!sig))

@@

\noindent where $d_i$ is the distance between $\text{stomate}_i$ and its nearest neighbor. We calculated $\mathrm{NNI}$ using the *R* package **spatialEco** version `r packageVersion("spatialEco")` [@evans_spatialeco_2023]. The observed stomatal distribution is dispersed relative to a uniform random distribution if the observed $NNI$ is greater than 95% of the synthetic $\mathrm{NNI}$ values (one-tailed test).

@@

<!-- WORKING HERE -->
Many leaf surface (`r nni[['n1']]` of `r nni[['n1']] + nni[['n2']]`, `r round(100 * nni[['n1']] / (nni[['n1']] + nni[['n2']]), 1)`%) are significantly overdispersed compared to a random uniform distribution, but none were close to an ideal hexagonal pattern (dispersion index = 1; \autoref{fig:single-surface}). Dispersion index differs significantly among light treatments (ANOVA, $F = `r$) is significantly greater Additionally, we hypothesized that as CO$_2$ is more limiting to photosynthesis under high light, the dispersion index would closer to 1 in high light...WORKING HERE would be more uniformly dispersed in plants grown in high light than plants grown in low and medium light. The data also fail to support this hypothesis as there is no strong, discernible trend between light and stomatal patterning. Interestingly, adaxial leaf surfaces were more uniformly dispersed than associated abaxial leaf surfaces across all light treatments (F$_\text{1,126}$ = 28.8; p-value < 0.001). Rather than regulate stomatal patterning in response to light regimes, plants respond by increasing stomatal density (\autoref{fig:density}). Stomatal density drastically increased in plants grown under high light, validating the long held hypothesis that light strongly influences stomatal density (F$_\text{2,126}$ = 680.7; p-value < 0.001).

>>>>

# Summarize results on NNI
nni1 = single_surface_results |>
  dplyr::transmute(sig = p_value < 0.05) |>
  dplyr::summarize(n1 = sum(sig), n2 = sum(!sig))

adjp = mt.rawp2adjp(single_surface_results$p_value, proc = "BH")

nni2 = bind_cols(single_surface_results[adjp$index,], adjp$adjp) |>
  dplyr::transmute(sig = BH < 0.05) |>
  dplyr::summarize(n1 = sum(sig), n2 = sum(!sig))

@@

\noindent where $d_i$ is the distance between $\text{stomate}_i$ and its nearest neighbor. We calculated $\mathrm{NNI}$ using the *R* package **spatialEco** version `r packageVersion("spatialEco")` [@evans_spatialeco_2023]. The observed stomatal distribution is dispersed relative to a uniform random distribution if the observed $NNI$ is greater than 95% of the synthetic $\mathrm{NNI}$ values (one-tailed test). We adjusted $P$-values to account for multiple comparisons using the Benjamini-Hochberg [@benjamini_controlling_1995] false discovery rate procedure implemented in the *R* package **multtest** version `r packageVersion("multtest")` [@pollard_multiple_2005].

@@

<!-- WORKING HERE -->
Many leaf surfaces (`r nni2[['n1']]` of `r nni2[['n1']] + nni2[['n2']]`, `r round(100 * nni2[['n1']] / (nni2[['n1']] + nni2[['n2']]), 1)`%) are significantly overdispersed compared to a random uniform distribution, but none were close to an ideal hexagonal pattern (dispersion index = 1; \autoref{fig:single-surface}). Before controlling for multiple comparisons, `r round(100 * nni1[['n1']] / (nni1[['n1']] + nni1[['n2']]), 1)`% are significantly overdispersed. The dispersion index differs significantly among light treatments (ANOVA, $F = `r$) is significantly greater Additionally, we hypothesized that as CO$_2$ is more limiting to photosynthesis under high light, the dispersion index would closer to 1 in high light...WORKING HERE would be more uniformly dispersed in plants grown in high light than plants grown in low and medium light. The data also fail to support this hypothesis as there is no strong, discernible trend between light and stomatal patterning. Interestingly, adaxial leaf surfaces were more uniformly dispersed than associated abaxial leaf surfaces across all light treatments (F$_\text{1,126}$ = 28.8; p-value < 0.001). Rather than regulate stomatal patterning in response to light regimes, plants respond by increasing stomatal density (\autoref{fig:density}). Stomatal density drastically increased in plants grown under high light, validating the long held hypothesis that light strongly influences stomatal density (F$_\text{2,126}$ = 680.7; p-value < 0.001).

### 2023-10-17 ms/ms.tex
added multtest results

\noindent where \(d_i\) is the distance between \(\text{stomate}_i\) and
its nearest neighbor. We calculated \(\mathrm{NNI}\) using the \emph{R}
package \textbf{spatialEco} version 2.0.1
(\protect\hyperlink{ref-evans_spatialeco_2023}{J. S. Evans and Murphy
2023}). The observed stomatal distribution is dispersed relative to a
uniform random distribution if the observed \(NNI\) is greater than 95\%
of the synthetic \(\mathrm{NNI}\) values (one-tailed test).

@@

To test whether the position of ab- and adaxial stomata are coordinated
we compared the observed distribution to a null distribution where the
positions on each surface are random. For each pair of surfaces
(observed or synthetic) we calculated the distance squared between each
to the nearest stomatal centroid with the \emph{R} package
\textbf{raster} version 3.6.23. Then we calculated the cell-wise Pearson
correlation coefficient. If stomatal positions on each surface are
coordinated to minimize the distance between mesophyll and the nearest
stomate, then we expect a negative correlation. A cell that is far from
a stomate on one surface should be near a stomate on the other surface
(\autoref{fig:ideal-amphi-grid}). We generated a null distribution of
the correlation coefficient by simulating \(10^3\) synthetic data sets
for each observed pair. For each synthetic data set, we simulated
stomatal position using a random uniform distribution, as described
above, matching the number of stomata on abaxial and adaxial leaf
surfaces. Stomatal positions on each surface are coordinated if the
correlation coefficient is greater than 95\% of the synthetic
correlation values (one-tailed test).

@@

Stomatal density of \emph{Arabidopsis thaliana} the 132 leaves measured
ranged from 12 to 93 (units) with high light leaves ranging from 93 to
55 (units), medium light from 15 to 35 (units), and low light from 12 to
42 (units). Leaves were amphistomatous with a mean stomatal ratio of
0.45.

@@

Many leaf surface (57 of 132, 43.2\%) are significantly overdispersed
compared to a random uniform distribution, none of the leaf surfaces
exhibited perfectly uniform stomatal patterning (dispersion index = 1)
(\autoref{fig:single-surface}). Additionally, we hypothesized that as
CO\(_2\) is more limiting to photosynthesis under high light, stomata
would be more uniformly dispersed in plants grown in high light than
plants grown in low and medium light. The data also fail to support this
hypothesis as there is no strong, discernible trend between light and
stomatal patterning. Interestingly, adaxial leaf surfaces were more
uniformly dispersed than associated abaxial leaf surfaces across all
light treatments (F\(_\text{1,126}\) = 28.8; p-value \textless{} 0.001).
Rather than regulate stomatal patterning in response to light regimes,
plants respond by increasing stomatal density (\autoref{fig:density}).
Stomatal density drastically increased in plants grown under high light,
validating the long held hypothesis that light strongly influences
stomatal density (F\(_\text{2,126}\) = 680.7; p-value \textless{}
0.001).

>>>>

\noindent where \(d_i\) is the distance between \(\text{stomate}_i\) and
its nearest neighbor. We calculated \(\mathrm{NNI}\) using the \emph{R}
package \textbf{spatialEco} version 2.0.1
(\protect\hyperlink{ref-evans_spatialeco_2023}{J. S. Evans and Murphy
2023}). The observed stomatal distribution is dispersed relative to a
uniform random distribution if the observed \(NNI\) is greater than 95\%
of the synthetic \(\mathrm{NNI}\) values (one-tailed test). We adjusted
\(P\)-values to account for multiple comparisons using the
Benjamini-Hochberg
(\protect\hyperlink{ref-benjamini_controlling_1995}{\textbf{benjamini\_controlling\_1995?}})
false discovery rate procedure implemented in the \emph{R} package
\textbf{multtest} version 2.56.0
(\protect\hyperlink{ref-pollard_multiple_2005}{\textbf{pollard\_multiple\_2005?}}).

@@

To test whether the position of ab- and adaxial stomata are coordinated
we compared the observed distribution to a null distribution where the
positions on each surface are random. For each pair of surfaces
(observed or synthetic) we calculated the distance squared between each
to the nearest stomatal centroid with the \emph{R} package
\textbf{raster} version 3.6.26. Then we calculated the cell-wise Pearson
correlation coefficient. If stomatal positions on each surface are
coordinated to minimize the distance between mesophyll and the nearest
stomate, then we expect a negative correlation. A cell that is far from
a stomate on one surface should be near a stomate on the other surface
(\autoref{fig:ideal-amphi-grid}). We generated a null distribution of
the correlation coefficient by simulating \(10^3\) synthetic data sets
for each observed pair. For each synthetic data set, we simulated
stomatal position using a random uniform distribution, as described
above, matching the number of stomata on abaxial and adaxial leaf
surfaces. Stomatal positions on each surface are coordinated if the
correlation coefficient is greater than 95\% of the synthetic
correlation values (one-tailed test).

@@

Stomatal density of \emph{Arabidopsis thaliana} the 132 leaf surfaces
measured ranged from 12 to 93 (units) with high light leaves ranging
from 93 to 55 (units), medium light from 15 to 35 (units), and low light
from 12 to 42 (units). Leaves were amphistomatous with a mean stomatal
ratio of 0.45.

@@

Many leaf surfaces (37 of 132, 28\%; ) are significantly overdispersed
compared to a random uniform distribution, but none were close to an
ideal hexagonal pattern (dispersion index = 1;
\autoref{fig:single-surface}). Before controlling for multiple
comparisons, 43.2\% are significantly overdispersed. The dispersion
index differs significantly among light treatments (ANOVA, \(F = `r\))
is significantly greater Additionally, we hypothesized that as CO\(_2\)
is more limiting to photosynthesis under high light, the dispersion
index would closer to 1 in high light\ldots WORKING HERE would be more
uniformly dispersed in plants grown in high light than plants grown in
low and medium light. The data also fail to support this hypothesis as
there is no strong, discernible trend between light and stomatal
patterning. Interestingly, adaxial leaf surfaces were more uniformly
dispersed than associated abaxial leaf surfaces across all light
treatments (F\(_\text{1,126}\) = 28.8; p-value \textless{} 0.001).
Rather than regulate stomatal patterning in response to light regimes,
plants respond by increasing stomatal density (\autoref{fig:density}).
Stomatal density drastically increased in plants grown under high light,
validating the long held hypothesis that light strongly influences
stomatal density (F\(_\text{2,126}\) = 680.7; p-value \textless{}
0.001).
